---
title: "Digital Phenotyping Mood Disorders Reproducible R Code"
author: "Nick Jacobson"
date: "2018-09-12T21:13:14-05:00"
output: html_document
categories: ["R"]
tags: ["Digital Phenotyping","Machine Learning", "Depression","Major depression","Bipolar","intensive-longitudinal-data", "lags"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Overview

This post documents reproducible code accompanying the manuscript draft "Digital Biomarkers of Mood Disorders and Symptom Change" by Nicholas C. Jacobson, Hilary M. Weingardenm and Sabine Wilhelm (published in Nature Partner Journal (npj): Digital Medicine). This code uses machine learning to predict the diagnostic status and depressive symptom change in a a group of 23 patients with bipolar disorder or major depressive disorder and 32 non-disordered controls using actigraphy data. 

The data in the following code is available publicly using the following [link](/http://datasets.simula.no/depresjon/data/depresjon-dataset.zip)

Read in the data.
```{r Load Mood Data}
setwd("C:/Users/njaco/Dropbox/Depresion and Motor Activity/data")
scores=read.csv("scores.csv")
```

Load the required packages.
```{r Load Packages}
library(psych) #for the rmssd function
library(e1071) #For distribution
library(mgcv) #To run stage 1 of DTVEM (see Jacobson, Chow, and Newman, 2018)
library(gtools) #For the mixsort
library(caret) #To run the machine learning techniques.
library(ggplot2) #For plots
library(scales) #For plots
library(Hmisc) #For correlations
```

Create and define biomarker new feature functions.
```{r Define functions}
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

meanvariabilitystats=function(x){
  rmssd1=rmssd(x,lag=1)
  rmssd2=rmssd(x,lag=2)
  max1=max(x,na.rm=TRUE)
  min1=min(x,na.rm=TRUE)
  mean1=mean(x,na.rm=TRUE)
  median1=median(x,na.rm=TRUE)
  mode1=getmode(x)
  sd1=sd(x,na.rm=TRUE)
  skewness1=skewness(x,na.rm=TRUE)
  kurtosis1=kurtosis(x,na.rm=TRUE)
  quantiles1=quantile(x,seq(.01,.99,by=.01))
  return(c(rmssd1,rmssd2,max1,min1,mean1,median1,mode1,sd1,skewness1,kurtosis1,quantiles1))
}

dtvemstage1=function(x){
  x2=data.frame(outcome=NA,lag=NA,pred=NA)
  for(i in 1:100){
    outcome=x[1:(length(x)-i)]
    pred=x[-1:-i]
    tempx=data.frame(outcome=outcome,lag=i,pred=pred)
    x2=rbind(x2,tempx)
  }
  out=bam(outcome~s(lag,by=pred),data=x2)
  pdat=data.frame(pred=1,lag=1:100)
  stage1est=predict(out,pdat,type="terms")
  return(stage1est)
}
```

Next we'll use the functions to create a set of digital biomarkers from the actigraphy data.
```{r Read in actigraphy data and create biomarkers}
setwd("C:/Users/njaco/Dropbox/Depresion and Motor Activity/data")

condlist=list.files("condition/")
condlist=mixedsort(condlist) #PUTS THEM IN NUMERICAL ORDER
conddatalist=list()
condspectrumlist=list()
condmeanvariabilitystats=list()
conddtvemstage1=list()

for(i in 1:length(condlist)){
  conddatalist[[i]]=read.csv(paste("condition/",condlist[i],sep=""))
  condspectrumlist[[i]]=stats::spectrum(conddatalist[[i]]$activity[1:19299]) #19299 is last minute available for all subjects
  condmeanvariabilitystats[[i]]=meanvariabilitystats(conddatalist[[i]]$activity)
  conddtvemstage1[[i]]=dtvemstage1(conddatalist[[i]]$activity)
}


controllist=list.files("control/")
library(gtools)
controllist=mixedsort(controllist) #PUTS THEM IN NUMERICAL ORDER
controldatalist=list()
controlspectrumlist=list()
controlmeanvariabilitystats=list()
controldtvemstage1=list()

for(i in 1:length(controllist)){
  controldatalist[[i]]=read.csv(paste("control/",controllist[i],sep=""))
  controlspectrumlist[[i]]=stats::spectrum(controldatalist[[i]]$activity[1:19299]) #19299 is last minute available for all subjects
  controlmeanvariabilitystats[[i]]=meanvariabilitystats(controldatalist[[i]]$activity)
  controldtvemstage1[[i]]=dtvemstage1(controldatalist[[i]]$activity)
}
```

Next merge the features into the main dataset.
```{r Merging data}
scores[,paste("f",1:9929,sep="")]=NA
#IGNORE SPECTRAL ANALSYSI FOR NOW
for(i in 1:length(condlist)){
  scores[i,paste("f",1:109,sep="")]=condmeanvariabilitystats[[i]]
  scores[i,paste("f",110:209,sep="")]=conddtvemstage1[[i]]
  scores[i,paste("f",210:9929,sep="")]=condspectrumlist[[i]]$spec
}

for(i in 1:length(controllist)){
  scores[(i+23),paste("f",1:109,sep="")]=controlmeanvariabilitystats[[i]]
  scores[(i+23),paste("f",110:209,sep="")]=controldtvemstage1[[i]]
  scores[(i+23),paste("f",210:9929,sep="")]=controlspectrumlist[[i]]$spec
}
```

#Result 1: Diagnostic Group Status

Define the outcome.
```{r Define outcome}
scores$TYPE=c(rep("Patient",length(condlist)),rep("Control",length(controllist)))
```

Create the training dataset for the mood classification. Note that this is important so that only the biomarker features are use in predicting the diagnostic status.
```{r Create the training set}
traindata=scores[,c("TYPE",paste("f",1:9929,sep=""))]
```

Train a machine learning model utilizing xgboost and leave-one-out cross-validation.
```{r Train the classification model}
train_control<- trainControl(method="LOOCV",savePredictions = TRUE)

set.seed(87626) #SET A RANDOM SEED
model<- train(TYPE~., data=traindata, trControl=train_control, method="xgbTree",na.action=na.pass) #xgboost
model
```

List the most important biomarkers.
```{r List Important Predictors}
varImp(model)
```

Next plot the classification results.
```{r Plot the classification results}
ggplotConfusionMatrix <- function(m){
  mytitle <- paste("Accuracy", percent_format()(m$overall[1]),
                   "Kappa", percent_format()(m$overall[2]))
  p <-
    ggplot(data = as.data.frame(m$table) ,
           aes(x = Reference, y = Prediction)) +
    geom_tile(aes(fill = log(Freq)), colour = "white") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    geom_text(aes(x = Reference, y = Prediction, label = Freq)) +
    theme(legend.position = "none") +
    ggtitle("Diagnostic Status",subtitle=mytitle)
  return(p)
}
xgpred=model$pred$pred[model$pred$eta==model$bestTune$eta&model$pred$max_depth==model$bestTune$max_depth&model$pred$gamma==model$bestTune$gamma&model$pred$colsample_bytree==model$bestTune$colsample_bytree&model$pred$min_child_weight==model$bestTune$min_child_weight&model$pred$subsample==model$bestTune$subsample&model$pred$nrounds==model$bestTune$nrounds]
observed=scores$TYPE
ggplotConfusionMatrix(caret::confusionMatrix(xgpred,observed))
```

Calculate the diagnostic group's results.
```{r Calculate the classification results}
caret::confusionMatrix(xgpred,observed)
```

#Result 2: Predicting Change in Depression
The next phase is to predict the change in depression across the two weeks.

Define the training set for the symptoms set.
```{r Define training set of the symptom data}
scores$changedep=scores$madrs2-scores$madrs1 #DEFINE CHANGE IN SYMPTOMS
traindataSX=scores[,c("changedep",paste("f",1:9929,sep=""))] #DEFINE TRAINING SET
traindataSX=traindataSX[!is.na(traindataSX$changedep),] #only use the patients.
```

Define a function that will maximize the correlation between variables. Note that the correlation function will result in a negative number so all predictions need to be reversed.
```{r}
cormax <- function(data, model = NULL,lev=NULL) {
  y_pred = data$pred
  y_true = data$obs
  cormax <- cor(y_pred,y_true)
  c(cormaxout=cormax)
}
```

Train the model.
```{r Train the symptom change model}
train_control<- trainControl(method="LOOCV",savePredictions = TRUE,summaryFunction = cormax)
set.seed(38552) #min(modelSX$results$cormaxout)=-0.7819468
modelSX<- train(changedep~., data=traindataSX, trControl=train_control, method="xgbTree",na.action=na.pass,maximize=FALSE,metric="cormaxout") #xgboost
modelSX 
```

Identify the predictions vs the observed values.
```{r Save symptom change predictions}

ps=modelSX$pred[modelSX$pred$eta==modelSX$bestTune$eta&modelSX$pred$max_depth==modelSX$bestTune$max_depth&modelSX$pred$gamma==modelSX$bestTune$gamma&modelSX$pred$colsample_bytree==modelSX$bestTune$colsample_bytree&modelSX$pred$min_child_weight==modelSX$bestTune$min_child_weight&modelSX$pred$subsample==modelSX$bestTune$subsample&modelSX$pred$nrounds==modelSX$bestTune$nrounds,c("pred","obs")]

ps[,1]=ps[,1]*-1 #REVERSE THE PREDICTIONS

```

Get the correlation between the predicted and observed symptom change.
```{r Calculate correlation}
cormat=rcorr(data.matrix(ps))
cormat$r[1,2] #CORRELATION
cormat$P[1,2] #P-VALUE
```

Plot the results.
```{r Plot the results for change in depressive symptoms}

longpreds=data.frame(Type=c(rep("Predicted",23),rep("Observed",23)),PatientIndex=c(1:23,1:23),zscore=c(scale(ps[,1]),scale(ps[,2])))

ggplot(data=longpreds,aes(x=PatientIndex, y=zscore, colour=Type)) +geom_line()+ylab("Z-Score")+xlab("Patient Index")+ggtitle(paste("Change in Depressive Symptoms"),subtitle=expression(paste(italic("r")," = 0.782, ",italic("p")," = 1.04e-05",sep="")))

```

